---
title: "Preparing your Environment"
output: html_notebook
---

The tidyverse is the name given to a diverse set of packages with a consistent "grammar". You have probably heard of some of them: **ggplot2**, **dplyr**, **readr**, to name a few. You can install _most_ of the tidyverse by installing the **tidyverse** package. We will omit the output here, because it is a little noisy! 

```{r, eval=FALSE}
install.packages("tidyverse")
```

With the **tidyverse** package installed, load it with `library()`. The **tidyverse** package loads (attaches) eight related packages:

1. **ggplot2**: for visualizing data
2. **tibble**: for working with table data
3. **tidyr**: for re-arranging and cleaning table data
4. **readr**: for reading in a diverse array of data types
5. **purrr**: for functional programming and mapping
6. **dplyr**: a powerful tool for data chunking and analysis
7. **stringr**: for working with character strings (e.g. words or codes)
8. **forcats**: for working with factors (e.g. categorical data)

You will see some output about function names that conflict with built-in functions. This is normal. 

```{r}
library(tidyverse)
```
